Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Whoopsie in redefining merge.
Bigger is better.

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\newcommand{\naive}{na\"ive}
\newcommand{\naively}{ na\"ively}
\newcommand{\Naive}{Na\"ive}
\newcommand{\Naively}{Na\"ively}
% \usepackage{xltxtra}
% \setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
%
\input{preamble}

@@

This article contributes a step by step process for going from a
\naive analyzer derived under the abstracting abstract machine
approach, to an efficient program analyzer.  The end result of the
process is a two to three order-of-magnitude improvement over the
systematically derived analyzer, making it competitive with
hand-optimized implementations that compute fundamentally less precise
results.
\end{abstract}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=3.2in]{church-relative-space.ps}
\includegraphics[width=3.2in]{church-relative-speed.ps}
\includegraphics[width=3.2in]{church-relative-time.ps}
\vspace{-1.5em}
\end{center}
\caption{Relative improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. Each point is marked with the
  section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

To conduct this \naive{} exploration on the \Church{} example would require
considerable time.  Even though the state space is finite, it is exponential in
the size of the program.  Even with $k = 0$, there are exponentially many
stores in the AAM framework.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentiality to merely
daunting polynomiality.

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it will install this joined store into current
state.

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in C\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all subsequent stores.

@@

\subsubsection{\Naively}

@@

\subsubsection{Less \naively{} and less precisely}

@@

\begin{figure}
\include{bench-overview}
\caption{Overview performance comparison between baseline and
  optimized analyzer.}
\label{fig:bench-overview}
\end{figure}

@@

[Computes results more like Earl, et al.'s 0-CFA, which times out on
  the Church numeral benchmark and so even though it offers a fair
  point of comparison, a more thorough evaluation is likely to
  uninformative as the other benchmarks are highly likely to timeout
  as well (and it would require significant effort to extend their
  implementation with the features needed to analyze our benchmark
  suite).  That implementation is evaluated against much smaller
  benchmarks: the largest program consists of 63 \emph{subexpressions}
  and times out after 30 minutes.]

Vardoulakis and Shivers evaluate their CFA2
analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0-CFA
defined in their framework.  [NEED TIMING FOR 0-CFA ON CHURCH.]  The
Church example is the largest benchmark Vardoulakis and Shivers
consider.  More work would be required to scale the analyzer to the
set of features required by our benchmarks.

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
% \usepackage{xltxtra}
% \setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
%
\input{preamble}

@@

This article contributes a step by step process for going from a naive
analyzer derived under the abstracting abstract machine approach, to
an efficient program analyzer.  The end result of the process is a two
to three order-of-magnitude improvement over the systematically
derived analyzer, making it competitive with hand-optimized
implementations that compute fundamentally less precise results.
\end{abstract}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=3.2in]{church-relative-space.ps}
\includegraphics[width=3.2in]{church-relative-speed.ps}
\includegraphics[width=3.2in]{church-relative-time.ps}
\vspace{-1.5em}
\end{center}
\caption{Relative improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where:
\begin{equation*}
\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\text.
\end{equation*}

To conduct this naive exploration on the \Church{} example would require
considerable time.  Even though the state space is finite, it is exponential in
the size of the program.  Even with $k = 0$, there are exponentially many
stores in the AAM framework.

@@

A common technique to accelerate convergence in flow analyses is to share
a common store that is monotonically updated as interpretation
continues. In the ISWIM language with only unary operators, this makes
our 0-CFA quartic in the size of the program, which in one step drops the
complexity from intractable exponentiality to merely daunting
polynomiality.

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. The
scaffolded reduction relation, $\widehat{\machstep}$, is as follows:

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in cs\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

@@

\subsubsection{Na\"ively}

@@

\subsubsection{Less na\"ively and less precisely}

@@

\begin{figure}
\begin{tabular}{@{}l|r|r|r|r|r|r|r@{}}
Program & LOC
& \multicolumn{2}{c|}{Time (ms)}
& \multicolumn{2}{c|}{Space (mb)}
& \multicolumn{2}{c@{}}{Speed (state/sec)}
\\
\hline\hline
nucleic &  3508 & 0 & 0 & 0 & 0 & 0 & 0\\
matrix  &  747  & ? & ? & ? & ? & ? & ?\\
nbody   & 1434  & ? & ? & ? & ? & ? & ?\\
earley  &  667  & ? & ? & ? & ? & ? & ?\\
maze    &  681  & ? & ? & ? & ? & ? & ?\\
church  &   41  & ? & ? & ? & ? & ? & ?\\
lattice &  214  & ? & ? & ? & ? & ? & ?\\
boyer   &  642  & ? & ? & ? & ? & ? & ?\\
mbrotz  &   68  & ? & ? & ? & ? & ? & ?
\end{tabular}
\caption{Overview performance comparison between baseline and
  optimized analyzer.}
\label{fig:bench-overview}
\end{figure}

@@

[Computes results more like Earl, et al.'s 0CFA, which times out on
  the Church numeral benchmark and so even though it offers a fair
  point of comparison, a more thorough evaluation is likely to
  uninformative as the other benchmarks are highly likely to timeout
  as well (and it would require significant effort to extend their
  implementation with the features needed to analyze our benchmark
  suite).  That implementation is evaluated against much smaller
  benchmarks: the largest program consists of 63 \emph{subexpressions}
  and times out after 30 minutes.]

Vardoulakis and Shivers evaluate their CFA2
analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0CFA
defined in their framework.  [NEED TIMING FOR 0CFA ON CHURCH.]  The
Church example is the largest benchmark Vardoulakis and Shivers
consider.  More work would be required to scale the analyzer to the
set of features required by our benchmarks.
